---
title: "SDS 322E: Project 1 Report"
output: html_document
---

## Names and EIDs of Group Members: Andy Park, AP59864



## Load the Data

```{r setup, include=FALSE}
## Do not modify this code chunk
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```



```{r, cache=FALSE, message=FALSE}
## This chunk may take a long time to run. Set 'cache=TRUE' in the chunk options
## to make it run faster in subsequent knittings.

championships <- read_tsv("WCA_export_championships.tsv.bz2")
competitions <- read_tsv("WCA_export_Competitions.tsv.bz2")
continents <- read_tsv("WCA_export_Continents.tsv.bz2")
countries <- read_tsv("WCA_export_Countries.tsv.bz2")
eligible_country_iso2s_for_championship <- read_tsv("WCA_export_eligible_country_iso2s_for_championship.tsv.bz2")
events <- read_tsv("WCA_export_Events.tsv.bz2")
formats <- read_tsv("WCA_export_Formats.tsv.bz2")
persons <- read_tsv("WCA_export_Persons.tsv.bz2")
ranksaverage <- read_tsv("WCA_export_RanksAverage_333.tsv.bz2")
rankssingle <- read_tsv("WCA_export_RanksSingle_333.tsv.bz2")
results <- read_tsv("WCA_export_Results_333.tsv.bz2")
roundtypes <- read_tsv("WCA_export_RoundTypes.tsv.bz2")
scrambles <- read_tsv("WCA_export_Scrambles.tsv.bz2")
```


## Questions
### Active Speed Cubers

How many active (3x3x3) speedcubers are there registered with the WCA? For this question an *active speeedcuber* is defined as any person registered in the WCA who has competed in at least two competitions in the years 2022--2024. 


```{r}
## Count the number of active speedcubers who competed in at least 2 competitions between 2022 and 2024.
competitions <- competitions %>%
  rename(competitionId = id) %>%
  mutate(competitionDate = as.Date(paste(year, month, day, sep = "-")))

activepartic <- results %>%
  filter(eventId == 333) %>%
  left_join(competitions %>% select(competitionId, competitionDate), by = "competitionId") %>%
  mutate(year = as.numeric(format(competitionDate, "%Y"))) %>%
  filter(year >= 2022, year <= 2024) %>%
  distinct(personId, competitionId)

activespeedsc <- activepartic %>%
  group_by(personId) %>%
  summarize(num_comp = n()) %>%
  filter(num_comp >= 2)

activespeedsc_count <- nrow(activespeedsc)
activespeedsc_count
```

****

### World Records

This question has two parts:

1. Who holds the current world record single? 

```{r}
## Identify the current world record holder for the 3x3x3 event.
current_wr <- rankssingle %>%
  filter(eventId == 333, worldRank == 1) %>%
  left_join(persons, by = c("personId" = "id")) %>%
  select(personId, name, best, worldRank) %>%
  mutate(best_seconds = best / 100)

current_wr
```

****


On what date was this record set?


```{r}
## Find the competition date when the current world record was set.
current_record <- rankssingle %>%
  filter(eventId == 333, worldRank == 1) %>%
  select(personId, best)
record_solve <- results %>%
  filter(eventId == 333,
         personId == current_record$personId,
         best == current_record$best) %>%  
  left_join(competitions %>% select(competitionId, competitionDate), by = "competitionId") %>%
  arrange(competitionDate) %>%  
  slice(1)
record_solve %>% select(competitionId, competitionDate)
current_record <- rankssingle %>%
  filter(eventId == 333, worldRank == 1) %>%
  select(personId, best)

record_solve <- results %>%
  filter(eventId == 333,
         personId == current_record$personId,
         best == current_record$best) %>%  
  left_join(competitions %>% select(competitionId, competitionDate), by = "competitionId") %>%
  arrange(competitionDate) %>%  
  slice(1)

record_solve %>% select(competitionId, competitionDate)

```

**Write your answer here.**



2. Who *previously* held the world record single? On what date was this previous record set?

```{r}
## Identify the previous world record holder and determine the competition date when the record was set.
previous_record <- rankssingle %>%
  filter(eventId == 333, worldRank == 2) %>%
  select(personId, best)

previous_record <- previous_record %>%
  left_join(persons, by = c("personId" = "id"))
previous_record_solve <- results %>%
  filter(eventId == 333,
         personId == previous_record$personId[1],
         best == previous_record$best[1]) %>%
  left_join(competitions %>% select(competitionId, competitionDate), by = "competitionId") %>%
  arrange(competitionDate) %>%  
  slice(1)

previous_record_solve %>% select(competitionId, competitionDate)

```

**Write your answer here.**


### Regional Rankings

This question has two parts:

1. Amongst all speedcubers, who is the top ranked male speedcuber (for best single solve) in Australia?

```{r}
## Identify the top ranked male speedcuber in Australia.
topm_australia <- rankssingle %>%
  filter(eventId == 333) %>% 
  left_join(persons %>% select(id, name, countryId, gender), 
            by = c("personId" = "id")) %>%
  left_join(countries %>% select(id, name) %>% rename(countryName = name), 
            by = c("countryId" = "id")) %>%
  filter(gender == "m", countryName == "Australia") %>% 
  arrange(best) %>% 
  slice(1)

topm_australia

```

**Write your answer here.**


2. Amongst all speedcubers, who is the top ranked female speedcuber (for best single solve time) in Europe?

```{r}
## Identify the top ranked female speedcuber in Europe.

topf_europe <- rankssingle %>%
  filter(eventId == 333) %>%
  left_join(persons %>% select(id, name, countryId, gender), 
            by = c("personId" = "id")) %>%
  left_join(countries %>% select(id, continentId, name) %>% rename(countryName = name), 
            by = c("countryId" = "id")) %>%
  filter(gender == "f", continentId == "_Europe") %>%
  arrange(best) %>%  
  slice(1)
topf_europe

```

**Write your answer here.**



### Time Until Sub-5

Having a time below 5 seconds is considered an elite achievement and most speedcubers have to complete a large number of solves before they can obtain a sub-5 second solve. 

**NOTE**: Each round of a competition has 5 solves that should be considered separately when counting the number of solves.


1. For the current top 10 speedcubers in the world (as recorded in the RanksSingle table), on average, how many solves did they have to do before achieving a sub-5 second solve?


```{r}
## Calculate the average number of solves needed for the top 10 speedcubers to achieve a sub 5 second solve.

top10_ids <- rankssingle %>%
  filter(eventId == 333, worldRank <= 10) %>%
  pull(personId)
solves_long <- results %>%
  filter(eventId == 333, personId %in% top10_ids) %>%
  left_join(competitions %>% select(competitionId, competitionDate), by = "competitionId") %>%
  pivot_longer(
    cols = starts_with("value"),
    names_to = "solve_in_round",
    values_to = "time"
  ) %>%
  mutate(solve_in_round = parse_number(solve_in_round)) %>%
  arrange(personId, competitionDate, solve_in_round)

first_sub5 <- solves_long %>%
  group_by(personId) %>%
  mutate(solve_number = row_number()) %>%
  filter(time < 500) %>%  
  slice(1) %>%           
  ungroup() %>%
  select(personId, solve_number)

avgsolv <- mean(first_sub5$solve_number)
avgsolv
```

**Write your answer here.**



2. For **one** of the top 10 speedcubers make a plot of their solve times vs. the date of the solve, with date on the x-axis and solve time on the y-axis. 


```{r}
## Plot the solve times over competition dates for one of the top 10 competitors.
top10_ids <- rankssingle %>%
  filter(eventId == 333, worldRank <= 10) %>%
  pull(personId)

selectedid <- top10_ids[1]

selectedres <- results %>%
  filter(eventId == 333, personId == selectedid) %>%
  left_join(competitions %>% select(competitionId, competitionDate), by = "competitionId") %>%
  pivot_longer(
    cols = starts_with("value"),
    names_to = "solve_number",
    values_to = "time"
  ) %>%
  mutate(time_seconds = time / 100) %>%
  arrange(competitionDate, solve_number)
ggplot(selectedres, aes(x = competitionDate, y = time_seconds)) +
  geom_line() +
  geom_point()
```


**Write your answer here.**


### Up-and-Coming Speed Cubers

Which speed cubers **not** in the top 10,000 (worldwide for single best time) should we keep an eye on for the near future? 

The idea here is to identify "up-and-coming" speedcubers who are not yet achieving elite times. Come up with a list of **five** speedcubers (provide their names and WCA IDs) that you have identified as "up-and-coming". There is no one way to answer this question and the goal is to provide an analysis of the data that justifies the selection of your five names.

```{r}
## Identify five up-and-coming speedcubers (not in the top 10,000) based on their single best times.
upandcoming <- rankssingle %>%
  filter(eventId == 333, worldRank > 10000) %>%
  arrange(best) %>%  
  slice_head(n = 5) %>% 
  left_join(persons %>% select(id, name), by = c("personId" = "id")) %>%
  select(personId, name, best, worldRank) %>%
  mutate(best_seconds = best / 100)

upandcoming
```

**Write your answer here.**



### Region Rivalries

Europe and North America are both regions with strong speedcubers in the WCA. 

Which region has the faster group of speedcubers on average? 

To answer this question, characterize each person using their best *average* score according to their listing in the `ranksaverage` table. In the `persons` table the `countryId` indicates each person's country affiliation. The `countries` table lists the region that each country is in via the `continentId` column (Europe is "_Europe" and North America is "_North America").

Before attempting to answer the question, state what you expect the answer to be below.

**I expect that European speedcubers might have a slightly faster average (lower) score than North American speedcubers.**



```{r}
## Compare the average best scores of European vs. North American speedcubers.
region_comp <- ranksaverage %>%
  left_join(persons %>% select(id, countryId), by = c("personId" = "id")) %>%
  left_join(countries %>% select(id, continentId), by = c("countryId" = "id")) %>%
  filter(continentId %in% c("_Europe", "_North America")) %>%
  group_by(continentId) %>%
  summarize(avg_best_seconds = mean(best) / 100)

region_comp

```

What do you conclude about speedcubers in Europe vs. North America?

**Write you answer here.**


### Alternative Explanations

Develop an alternative explanation/hypothesis regarding speedcubers from Europe and North America that is 

1. Consistent with the results you produced in the previous question; but

2. Provides a different interpretation or explanation for what is going on. 

If the results from the previous question were unexpected, make use of systems thinking to develop an alternative hypothesis. If the results were consistent with your expectations, then use skeptical thinking. In either case, you should present an analysis that shows evidence for or against this alternative explanation relative to the conclusion that you made in the previous question.

**The alternative explanation is that the observed performance differences between European and North American speedcubers may be driven not by differences in intrinsic ability or training techniques, but rather by how often they compete. In this view, European speedcubers might be attending a larger number of competitions, which in turn provides them with more practice, exposure to competitive pressure, and opportunities for rapid feedback and learning. This higher competition frequency could lead to incremental improvements in solve times, creating a performance advantage over North American speedcubers who might participate less frequently. In our analysis, we merged the competition results with person and country data and then summarized the number of unique competitions attended by each competitor in Europe and North America. If the summary statistics indicate that European competitors have a higher average or median number of competitions, this supports the hypothesis that increased competitive experience is a significant factor in achieving faster solve times. Conversely, if the competition frequency is similar across regions, then factors other than competition exposure might be at play. Thus, while the original conclusion might attribute the performance gap to factors like training or innate skill, this alternative hypothesis emphasizes the role of competition frequency; a system-level factor that can enhance learning and performance over time.**

```{r}
## Evaluate an alternative hypothesis that differences in competition frequency might explain the regional differences.
compfreq <- results %>%
  left_join(persons %>% select(id, countryId), by = c("personId" = "id")) %>%
  left_join(countries %>% select(id, continentId), by = c("countryId" = "id")) %>%
  filter(continentId %in% c("_Europe", "_North America")) %>%
  group_by(personId, continentId) %>%
  summarize(num_competitions = n_distinct(competitionId), .groups = "drop")
compfreqsummary <- compfreq %>%
  group_by(continentId) %>%
  summarize(avg_competitions = mean(num_competitions),
            median_competitions = median(num_competitions),
            sd_competitions = sd(num_competitions), .groups = "drop")
compfreqsummary
```

**Summarize the results of your alternative analysis here.**





## Discussion

